% R Bootcamp, Module 8: Graphics
% August 2015, UC Berkeley
% Chris Krogslund (ckrogslund@berkeley.edu)

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(plyr)
library(reshape2)
library(ggplot2)
library(lattice)
library(gridExtra)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
data <- read.csv("../data/cpds.csv")
```

# By way of introduction...

* 3 main facilities for producing graphics in R: **base**, **lattice**, and **ggplot2**
  * In practice, these facilities are grouped into two camps: "basic" and "advanced"
* A better formulation: quick/dirty v. involved/fancy

# quick/dirty v. involved/fancy

* Recall that R is an object-oriented programming language

```{r}
tips<-reshape2::tips # Load dataset on tipping behavior included with reshape2 package
attributes(tips) # Check attributes of the tips dataset (names, row.names, class)
# Create an object of class "lm" (linear model), regressing tip on some covariates
tips.reg<-lm(formula=tip~total_bill+sex+smoker+day+time+size, data=tips)
attributes(tips.reg) # Check attributes of the tips.reg object (names, class)
```

* Base graphics often recognizes the object type and will implement specific plot methods

```{r, fig.cap=" "}
plot(tips) # Calls plotting method for class of tips dataset ("data.frame")
plot(tips.reg, which=1:2) # Calls plotting method for class of tips.reg objects ("lm"), print first two plots only
```

* lattice and ggplot2 generally **don't** exhibit this sort of behavior
```{r, fig.cap=" "}
xyplot(tips) # Attempt in lattice to automatically plot objects of class "data.frame"
ggplot(data=tips)+geom_point() # Attempt in ggplot to automatically plot objects of class "data.frame"
xyplot(tips.reg) # Attempt in lattice to automatically plot objects of class "lm"
ggplot(data=tips.reg)+geom_point() # Attempt in ggplot to automatically plot objects of class "lm"
```

* Easiest to cover **lattice** and **ggplot2** in tandem 

# The Dataset

* **Comparative Political Data Set I** (Armingeon et al. 2012)
* Cases: 23 industrialized democracies, 1960-2012
* Variables: *Government composition* (L-R); *state structure* (federalism, presidentialism, bicameralism, etc.); *macroeconomic indicators* (output, inflation, unemployment, deficit/debt, etc.); *demographics* (population, elderly)

For more info: [http://www.ipw.unibe.ch/content/team/klaus_armingeon/comparative_political_data_sets/index_eng.html](http://www.ipw.unibe.ch/content/team/klaus_armingeon/comparative_political_data_sets/index_eng.html)

# lattice v. ggplot2

* lattice is 

a) faster (though only noticeable over many and large plots) 

b) simpler (at first)

c) better at trellis graphs

d) able to do 3d graphs

* ggplot2 is 

a) generally more elegant

b) more syntactically logical (and therefore simpler, once you learn it)

c) better at grouping

d) able to interface with maps

# Basic usage: lattice

The general call for lattice graphics looks something like this:
  
```{r, eval=FALSE}
graph_type(formula, data=, [options])
```

The specifics of the **formula** differ for each graph type, but the general format is straightforward

```{r, eval=FALSE}
y             # Show the distribution of y

y~x           # Show the relationship between x and y 

y~x|A         # Show the relationship between x and y conditional on the values of A

y~x|A*B       # Show the relationship between x and y conditional on the combinations of A and B

z~y*x         # Show the 3D relationship between x, y, and z
```

# Basic usage: ggplot2

The general call for ggplot2 graphics looks something like this:
  
```{r, eval=FALSE}
ggplot(data=, aes(x=,y=, [options]))+geom_xxxx()+...+...+...
```

Note that ggplot2 graphs in layers in a *continuing call* (hence the endless +...+...+...), which really makes the extra layer part of the call

```{r, eval=FALSE}
...+geom_xxxx(data=, aes(x=,y=,[options]),[options])+...+...+...
```
You can see the layering effect by comparing the same graph with different colors for each layer

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=realgdpgr))+geom_point(color="black")+geom_point(aes(x=year, y=unemp), color="red")
ggplot(data=data, aes(x=year, y=realgdpgr))+geom_point(color="red")+geom_point(aes(x=year, y=unemp), color="black")
```

# Comparing lattice and ggplot

* Density Plots
* Scatter Plots
* Line Plots
* Bar plots
* Box plots
* Trellis Plots
* Contour Plots
* Tile/Image Plots
* 3D Plots (lattice)
* Panel Plots (ggplot2)

# lattice v. ggplot2: Densities

```{r, fig.cap=" ", warning=FALSE}
densityplot(~vturn, data=data) # lattice
ggplot(data=data, aes(x=vturn))+geom_density() # ggplot2
```

# lattice v. ggplot2: X-Y scatter plots

```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data) # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point() # ggplot2
```

# lattice v. ggplot2: X-Y line plots

```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data[data$country=="USA",], type="l") # lattice
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line() # ggplot2 
```

# lattice v. ggplot2: bar plots 
```{r, fig.cap=" ", warning=FALSE, message=FALSE}
# Create data.frame of average growth rates by country over time
growth<-ddply(.data=data, .variables=.(country), summarize, mean=mean(realgdpgr, na.rm=T))

barchart(mean~country, data=growth) # lattice
ggplot(data=growth, aes(x=country, y=mean))+geom_bar(stat="identity") # ggplot2
```

# lattice v. ggplot2: box plots 
```{r, fig.cap=" ", warning=FALSE}
bwplot(outlays~country, data=data) # lattice
ggplot(data=data, aes(x=country, y=outlays))+geom_boxplot() # ggplot2
```

# lattice v. ggplot2: "trellis" plots 
```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year|country, data=data) # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point()+facet_wrap(~country) # ggplot2
```

# lattice v. ggplot2: countour plots
```{r, fig.cap=" ", warning=FALSE}
data(volcano) # Load volcano contour data
volcano[1:10, 1:10] # Examine volcano dataset (first 10 rows and columns)
volcano3d <- melt(volcano) # Use reshape2 package to melt the data
head(volcano3d) # Examine volcano3d dataset (head)
names(volcano3d) <- c("xvar", "yvar", "zvar") # Rename volcano3d columns
contourplot(zvar~xvar+yvar, data=volcano3d) # lattice
ggplot(data=volcano3d, aes(x=xvar, y=yvar, z = zvar))+geom_contour() # ggplot2
```

# lattice v. ggplot2: tile/image/level plots
```{r, fig.cap=" ", warning=FALSE}
levelplot(zvar~xvar+yvar, data=volcano3d) # lattice
ggplot(data=volcano3d, aes(x=xvar, y=yvar, z = zvar))+geom_tile(aes(fill=zvar)) # ggplot2
```

# lattice: 3D plots
```{r, fig.cap=" ", warning=FALSE}
# Create a subset of the dataset containing only data for France
france.data<-data[data$country=="France",]
cloud(outlays~year*realgdpgr, data=france.data)
# Create a subset of the dataset containing only data for Greece, Portugal, Ireland, and Spain
pigs.data<-data[data$country %in% c("Greece", "Portugal", "Ireland", "Spain"),]
cloud(outlays~year*realgdpgr|country, data=pigs.data)
```

# ggplot2: Panel plots
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=pigs.data, aes(x=year, y=realgdpgr, color=country))+geom_line()
```

# lattice v. ggplot2: options [labeling]

```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data, xlab="Year", ylab="Government Outlays", main="Cool Graph") # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point()+xlab(label="Voter Turnout (%)")+ylab(label="Government Outlays")+ggtitle(label="Cool Graph") # ggplot2
```

# lattice v. ggplot2: options [axis + size scaling]

```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data) # lattice
xyplot(outlays~year, data=data, cex=2) # lattice
xyplot(outlays~year, data=data, cex=.5) # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point() # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(size=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(size=1) # ggplot2
```

# lattice v. ggplot2: options [graphical parameters]

* Colors
```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data, col=colors()[145]) #lattice
xyplot(outlays~year, data=data, col="red") #lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point(color=colors()[145]) # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(color="red") # ggplot2
```

* Point Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data, pch=3) # lattice
xyplot(outlays~year, data=data, pch=15) # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape=15) # ggplot2
```

* Point Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data, pch=3) # lattice
xyplot(outlays~year, data=data, pch=15) # lattice
xyplot(outlays~year, data=data, pch="w") # lattice
xyplot(outlays~year, data=data, pch="$", cex=2) # lattice
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape=15) # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape="w") # ggplot2
ggplot(data=data, aes(x=year, y=outlays))+geom_point(shape="$", size=5) # ggplot2
```

* Line Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=1) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=2) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=2) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=3) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=4) # lattice
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=1) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=2) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=3) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=3, size=1) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=3, size=1.5) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays))+geom_line(linetype=3, size=2) # ggplot2
```

# ggplot2 and the Grammar of Graphics

- By now, you might be noticing some trends in how these two packages approach graphics

- lattice tends to focus on a particular type of graph and how to represent cross-sectional variation by splitting it up into smaller chunks

- Becoming a proficient user of lattice requires learning a huge array of graph-specific formulas and options

- ggplot2 tries to represent much more of the cross-sectional variation by making use of various "aesthetics"; general approach is based on *The Grammar of Graphics*
  
# ggplot2 and the Grammar of Graphics
  
  - Basic idea is that the visualization of all data requires four items

1) One or more **statistics** conveying information about the data (identities, means, medians, etc.)

2) A **coordinate system** that differentiates between the intersections of statistics (at most two for ggplot, three for lattice)

3) **Geometries** that differentiate between off-coordinate variation in *kind*
  
  4) **Scales** that differentiate between off-coordinate variation in *degree*
  
  - ggplot2 allows the user to manipulate all four of these items

# Anatomy of aes()

```{r, eval=FALSE}
ggplot(data=, aes(x=, y=, color=, linetype=, shape=, size=))
```

ggplot2 is optimized for showing variation on all four aesthetic types

```{r, fig.cap=" ", warning=FALSE}
# Differences in kind using color
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_line(aes(color=country))
```

Note what happens when we specify the color parameter outside of the aesthetic operator. ggplot2 views these specifications as invalid graphical parameters. 

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_line(color=country)
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_line(color="country")
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_line(color="red")
```

```{r, fig.cap=" ", warning=FALSE}
# Differences in kind using line types
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_line(aes(linetype=country))
# Differences in kind using point shapes
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(shape=country))
# Differences in degree using color
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(color=realgdpgr))
# Differences in degree using point size
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(size=realgdpgr))
# Multiple non-cartesian aesthetics (differences in kind using color, degree using point size)
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(color=country,size=realgdpgr))
```

# Fitted lines and curves with ggplot2
```{r, fig.cap=" ", warning=FALSE, message=FALSE}
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()
# Add linear model (lm) smoother
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()+geom_smooth(method="lm")
# Add local linear model (loess) smoother, span of 0.75 
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()+geom_smooth(method="loess", span=.75)
# Add local linear model (loess) smoother, span of 0.25 
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()+geom_smooth(method="loess", span=.25)
# Add linear model (lm) smoother, no standard error shading 
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()+geom_smooth(method="lm", se=F)
# Add local linear model (loess) smoother, no standard error shading 
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point()+geom_smooth(method="loess", se=F)
# Add a local linear (loess) smoother for each country
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(color=country))+geom_smooth(aes(color=country))
# Add a local linear (loess) smoother for each country, no standard error shading
ggplot(data=pigs.data, aes(x=year, y=outlays))+geom_point(aes(color=country, size=realgdpgr))+geom_smooth(aes(color=country), se=F)
```

# lattice v. ggplot2: tables

* Both lattice and ggplot2 graphs can be combined using the *grid.arrange()* function in the **gridExtra** package 
```{r, warning=FALSE, fig.cap=" "}
# Initialize gridExtra library
library(gridExtra)
# Create 3 plots to combine in a table
plot1<-ggplot(data=pigs.data, aes(x=year, y=outlays, color=))+geom_line(aes(color=country))
plot2<-ggplot(data=pigs.data, aes(x=year, y=outlays, linetype=))+geom_line(aes(linetype=country))
plot3<-ggplot(data=pigs.data, aes(x=year, y=outlays, shape=))+geom_point(aes(shape=country))
# Call grid.arrange
grid.arrange(plot1, plot2, plot3, nrow=3, ncol=1)
```

# Exporting

Two basic image types

1) **Raster/Bitmap** (.png, .jpeg)

Every pixel of a plot contains its own separate coding; not so great if you want to resize the image

```{r, eval=FALSE}
jpeg(filename="example.png", width=, height=)
plot(x,y)
dev.off()
```

2) **Vector** (.pdf, .ps)

Every element of a plot is encoded with a function that gives its coding conditional on several factors; great for resizing

```{r, eval=FALSE}
pdf(filename="example.pdf", width=, height=)
plot(x,y)
dev.off()
```

# Exporting with lattice v. ggplot

```{r, eval=FALSE}
# Assume we saved our plot is an object called example.plot

# lattice
trellis.device(device="pdf", filename="example.pdf")
print(example.plot)
dev.off()

# ggplot2
ggsave(filename="example.pdf", plot=example.plot, scale=, width=, height=) # ggplot2
```

# Breakout

You're welcome to try out either lattice or ggplot for these questions, but in the solutions we'll focus on the ggplot approach.

For some of these you may want to use a smaller version of the dataset, such as a random subset, `subset <- air[sample(1:nrow(air), 10000, replace = FALSE), ]`.

### Basics

1) Plot a histogram of the flight delays with negative delays set to zero, censoring delay times at a maximum of 60 minutes.

2) Plot the arrival delay against the departure delay as a scatterplot.

3) Clean up your scatterplot with a title and axis labels

4) Make a boxplot of the departure delay as a function of the day of week.

### Using the ideas

5) Monkey with the size of points and style of points in your scatterplot and export to a pdf. Try to get it so that you'd be comfortable putting your plot ina report/paper.

6) Create a trellis plot of departure delay boxplots, one per destination for this subset of destinations, `DestSubset <- c('LAX','SEA','PHX','DEN','MSP','JFK','ATL','DFW','IAH', 'ORD')`.

7) Subset the data to flights going to Chicago (ORD) and Houston (IAH). Plot delay against scheduled departure time (CRSDepTime). Now plot so that flights to Chicago are in on color and  those to Houston in another. Use `scale_x_continuous()` and `scale_y_continuous()` to set the x-axis limits to be in the range from 6 am to midnight and the y-axis limits to be in the range (-10, 120).


### Advanced 

8) Try to create a single plot that shows the smoothed densities of departure delay with a different color or line type for each of the subset of destinations above. See if this allows you to better understand the effect of destination on delay. 

9) Using ggplot2, create a trellis plot where, for a given destination, each panel uses a) HOLLOW CIRCLES to plot departure delay as a function of time of day, and b) a red LOESS smoother without standard errors to plot the trend in departure delay over time of day.  BONUS: limit the time of day shown to 6 am to midnight, and turn off the grey background.


